Further Evidence of Modified Spin-down in Sun-like Stars:\\Pileups in the Temperature--Period Distribution

### 2022-04-18 src/ms.tex
coauthor comments

\subsection{Where do the pileups end?} \label{subsec:extent}
It appears from Figures~\ref{fig:kde} and \ref{fig:models} that the number density of stars on the \edit1{long-period pileup} declines towards cooler \teff, as predicted by the WMB model (see Figure~13 of \jvs). \edit1{The number density of stars on the short-period pileup similarly declines towards cooler \teff. If the short-period pileup is due to core–envelope coupling, one might expect an opposite trend of increasing number density toward cooler \teff, since the core–envelope coupling timescales and hence the ``stalled'' braking phases are longer for lower-mass stars \citep{Curtis2020}. However, the observed \teff--\prot distribution will also depend sensitively on the selection functions and observational biases inherent to both Kepler and the source of \teff (e.g. LAMOST).}

@@

\subsection{Detection biases}
\label{subsec:detectionbias}
\edit1{To this point, we have not addressed the detection biases inherent to Kepler rotation period catalogs. As periods become longer, they become more difficult to measure directly from Kepler time series. This is true for at least three reasons: (1) the observational baseline is finite, (2) photometric variation amplitude declines with increasing rotation period, and (3) photometric variations tend to become less periodic with increasing rotation period. The last two points are due to astrophysical reasons. This is important because, if a period detection threshold is Ro-dependent and the detection threshold Ro$_\mathrm{thresh}$ is close to the value \rocrit at which WMB becomes important, then the location of the pileup in the catalogs considered here depends sensitively on how detectability depends on Ro. For example, if Ro$_\mathrm{thresh}$ is only slightly larger than \rocrit and the long-period pileup has some non-negligible width in period space (due either to astrophysics or measurement error), then it's possible the current data set only reveals the lower edge of the long-period pileup.}

\edit1{\jvs addressed detection biases in detail, finding that a threshold of Ro$_\mathrm{thresh} \approx 2$ could reproduce the observed upper edge of the \mma \teff–\prot distribution. In this scenario, if \rothresh $\lesssim$ \rocrit, it becomes difficult to distinguish between the standard and WMB models in the \mma sample. However, those authors presented arguments that suggest detection bias is not solely responsible for producing the observed \teff–\prot distribution. First, stars at the upper edge are not those with the lowest-amplitude signals detected, as one would expect from a detection edge. For the variability amplitudes typical of Sun-like stars at the upper edge, one finds cooler stars with similar amplitude at even longer periods, as well as hotter stars with much smaller amplitudes (though at shorter periods). Even among Sun-like stars, those authors found that 9\% of stars below the edge have amplitudes smaller than stars with similar \teff at the edge (possibly due to issues such as flux dilution).}

\edit1{To understand whether the period detections on the long-period pileup are reliable, we examined the weight parameters published by \mma. The weight, $w$, is a metric \mma designed to serve as a proxy for the reliability of a rotation period detection, based on a star's autocorrelation function (ACF) properties and its position in $T_\mathrm{eff}$–$P_\mathrm{rot}$–LPH space, where LPH designates the local peak height of the ACF. Those authors selected $w>0.25$ for an acceptable compromise between real detections and false positives. In Figure~\ref{fig:weights}, we show that the long-period pileup is apparent even for $w>0.3$, and we conclude that the majority of stars on the long-period pileup are likely to be genuine detections and not an artifact of some detection threshold. Critically, we also note that while an Ro-dependent detection threshold may reproduce the upper edge of the \teff–\prot distribution, it would not be responsible for producing a pileup or overdensity at this edge. To our knowledge, there are no systematics of the rotation period catalogs considered here that would bias stars towards the detection threshold.}

@@

\edit1{We do note, however, that we can not rule out the possibility that the long-period pileup is wider in period or Rossby space than we observe, with stars just above the pileup being undetected through rotational brightness modulations. Such a scenario might result if magnetic braking is gradually weakened as stars approach \rocrit, as opposed to ceasing entirely at \rocrit. However, if the pileup does have a broader width in Rossby number, the close agreement between our observations and the \hall and \masuda samples suggests the true pileup width cannot be much larger than we observe.}

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{./figures/weights.pdf}
    \caption{The \teff–\prot distribution of the \lamostmcq sample for different weight thresholds (indicated above each panel) computed by \mma.}
    \label{fig:weights}    
\end{figure*}

@@

\begin{enumerate}
    \item \edit1{The \taucz relation adopted impacts the inferred \rocrit. We explored the \taucz formulae of \citet{BarnesKim2010}, \citet{Landin2010}, and \citet{Amard2019}, but these produce larger \taucz at fixed \teff than the \citet{CranmerSaar2011} relation, leading to even smaller inferred \rocrit values. We also explored using a fit to the \taucz computed in \jvs for solar-metallicity stars, which produces a smaller \taucz at fixed \teff than the \citet{CranmerSaar2011} relation. This relation revises our \rocrit estimates upwards by 15\%.} 
    \item \edit1{While our \rocrit inference procedure includes a constant \teff offset between the data and the \taucz relation, it does not include any \teff-dependent offsets. In Appendix~\ref{app:teff}, we show that the LAMOST \teff appear to have a non-linear mapping to the other \teff scales considered here. Perhaps not coincidentally, the \lamostmcq sample favors a \rocrit that is the least compatible with other samples. We also note that the MCMC results for the \lamostmcq sample preferred the LAMOST \teff to be shifted lower (Table~\ref{tab:mcmc}), which is in inconsistent with our finding that the LAMOST \teff are already lower than those from other spectroscopic surveys (Appendix~\ref{app:teff}).} 
    \item \edit1{Defining the long-period pileup or edge is a choice that impacts \rocrit. For example, though we chose the 90th \prot percentile in the \lamostmcq sample, \jvs chose the 95th percentile. Similarly, our boundary for the CKS long-period pileup was chosen subjectively.}
\end{enumerate}

@@

\edit1{Finally, we note that we regard the short-period pileup as less secure than the long-period pileup. In Appendix~\ref{app:harmonics} we show that the short-period pileup does not appear to be a harmonic of the long-period pileup, as one might expect if the periods on the short-period pileup were erroneously determined. However, the short-period pileup is primarily observed in the \lamostmcq sample, while it is unclear whether it is present in the \lamostsan sample (see the rightmost panels of Figure~\ref{fig:xmatch}). The differences between the \mma and \santos catalogs may provide an explanation for this observation. The latter catalog contains a much larger number of detections at longer rotation periods, and the higher sensitivity to more slowly rotating stars within the \santos may then make the short-period pileup appear weaker relative to the long-period pileup. This is as expected since open cluster observations suggest that the short-period pileup can not be a long-lived feature, while the long-period pileup appears to be much longer lived. Consequently, the \santos catalog may more accurately reflect the relative strengths of these two features.}

\edit1{An alternative explanation for the short-period overdensity is contamination by subgiants. Stars which have \teff on the ZAMS higher than $\sim$6250 will never undergo magnetic braking, spinning rapidly for the entirety of their main-sequence lifetimes. As these stars evolve away from the main-sequence, their temperatures drop while \logg changes minimally, while the spin rate slows due to radial expansion. While the feature persists even after crude subgiant cuts, it is possible a more careful selection would weaken the feature. We leave this to future investigators.}

@@

\edit1{If the Sun is not on the long-period pileup, this would seem to imply any number of unusual explanations. For example, the Sun's angular momentum evolution may have been different from those stars on the pileup. Alternatively, spin-down may proceed on the main-sequence after the pileup either through magnetic braking or some other mechanism. We consider these explanations unlikely, and suggest it is likely some combination of \teff systematics and observational biases that complicate contextualizing the Sun.}

@@

    \item It is yet unclear whether the Sun resides on the long-period pileup or has already evolved through it. \edit1{Offsets of $\approx$110--140~K between the observational and theoretical \teff scales allow for the long-period pileup to be fit with a solar value Rossby curve, thereby placing the Sun on the pileup.} If the Sun has evolved through the pileup, there is some modest tension with the Sun's age and the ages of the oldest stars on the pileup; some of the cooler long-period pileup stars in the CKS sample are both more massive and older than the Sun, which contradicts the expectation from the WMB model that hotter stars spend a shorter period of time on the pileup. However, this tension might be simply explained by inaccurate isochrone ages.

>>>>

\subsection{Where do the pileups end?} \label{subsec:extent}
It appears from Figures~\ref{fig:kde} and \ref{fig:models} that the number density of stars on the \edit1{long-period pileup} declines towards cooler \teff, as predicted by the WMB model (see Figure~13 of \jvs). \edit1{The number density of stars on the short-period pileup similarly declines towards cooler \teff. If the short-period pileup is due to core–envelope coupling, one might expect an opposite trend of increasing number density toward cooler \teff, since the core–envelope coupling timescales and hence the ``stalled'' braking phases are longer for lower-mass stars \citep{Curtis2020}. However, the observed \teff--\prot distribution depends sensitively on the selection functions and observational biases inherent to both Kepler and the source of \teff (e.g. LAMOST), which likely explains the observed number density trend.}

@@

\subsection{Detection biases}
\label{subsec:detectionbias}
\edit1{To this point, we have not addressed the detection biases inherent to Kepler rotation period catalogs. As periods become longer, they become more difficult to measure directly from Kepler time series. This is true for at least three reasons: (1) photometric variation amplitude declines with increasing Rossby number, (2) the existence of significant spot groups appears to be intermittent for high Rossby numbers, and (3) the observational baseline is finite. We note that the last point is not so important for Kepler's long baseline ($\sim$1400~d), while the first two points are due to astrophysical reasons. This is important because, if a period detection threshold is Ro-dependent and the detection threshold Ro$_\mathrm{thresh}$ is close to the value \rocrit at which WMB becomes important, then the location of the pileup in the catalogs considered here depends sensitively on how detectability depends on Ro. For example, if Ro$_\mathrm{thresh}$ is only slightly larger than \rocrit and the long-period pileup has some non-negligible width in period space (due either to astrophysics or measurement error), then it's possible the current data set only reveals the lower edge of the long-period pileup.}

\edit1{\jvs addressed detection biases in detail, finding that a threshold of Ro$_\mathrm{thresh} \approx 2$ could reproduce the observed upper edge of the \mma \teff–\prot distribution. In this scenario, if \rothresh $\lesssim$ \rocrit, it becomes difficult to distinguish between the standard and WMB models in the \mma sample. However, those authors presented arguments that suggest detection bias is not solely responsible for producing the observed \teff–\prot distribution. We direct readers to \S4.2 of that work for a summary of those arguments. Critically, we note here that while an Ro-dependent detection threshold may reproduce the upper edge of the \teff–\prot distribution, it would not be responsible for producing a pileup or overdensity at this edge. To our knowledge, there are no systematics of the rotation period catalogs considered here that would bias stars towards the detection threshold.}

\edit1{To understand whether the period detections on the long-period pileup are reliable, we examined the weight parameters published by \mma. The weight, $w$, is a metric \mma designed to serve as a proxy for the reliability of a rotation period detection, based on a star's autocorrelation function (ACF) properties and its position in $T_\mathrm{eff}$–$P_\mathrm{rot}$–LPH space, where LPH designates the local peak height of the ACF. Those authors selected $w>0.25$ for an acceptable compromise between real detections and false positives. In Figure~\ref{fig:weights}, we show that the long-period pileup is apparent even for $w>0.3$, and we conclude that the majority of stars on the long-period pileup are likely to be genuine detections and not an artifact of some detection threshold.}

@@

\edit1{We do note, however, that we can not rule out the possibility that the long-period pileup is wider in period or Rossby space than we observe, with stars just above the pileup being undetected through rotational brightness modulations. Such a scenario might result if magnetic braking is gradually weakened once stars reach \rocrit, as opposed to ceasing entirely at \rocrit. However, even in this scenario there are observational constraints on the rate at which the braking index changes. If the braking index starts to decline as Ro approaches \rocrit then there may be tension with the open cluster data. Additionally, if the pileup does have a broader width in Rossby number, the close agreement between our observations and the \hall and \masuda samples suggests the true pileup width cannot be much larger than we observe.}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{./figures/weights.pdf}
    \caption{The \teff–\prot distribution of the \lamostmcq sample for different weight thresholds (indicated above each panel) computed by \mma.}
    \label{fig:weights}    
\end{figure}

@@

\begin{enumerate}
    \item \edit1{The \taucz relation adopted impacts the inferred \rocrit. We explored the \taucz formulae of \citet{BarnesKim2010}, \citet{Landin2010}, and \citet{Amard2019}, but these produce larger \taucz at fixed \teff than the \citet{CranmerSaar2011} relation, leading to even smaller inferred \rocrit values. We also explored using a fit to the \taucz computed in \jvs for solar-metallicity stars, which produces a smaller \taucz at fixed \teff than the \citet{CranmerSaar2011} relation. This relation revises our \rocrit estimates upwards by 15\%.} 
    \item \edit1{While our \rocrit inference procedure includes a constant \teff offset between the data and the \taucz relation, it does not include any \teff-dependent offsets. In Appendix~\ref{app:teff}, we show that the LAMOST \teff appear to have a non-linear mapping to the other \teff scales considered here. Perhaps not coincidentally, the \lamostmcq sample favors a \rocrit that is the least compatible with other samples. We also note that the MCMC results for the \lamostmcq sample preferred the LAMOST \teff to be shifted lower (Table~\ref{tab:mcmc}), which is in inconsistent with our finding that the LAMOST \teff are already lower than those from other spectroscopic surveys (Appendix~\ref{app:teff}).} 
    \item \edit1{Defining the long-period pileup or edge is a choice that impacts the inferred value of \rocrit. For example, though we chose the 90th \prot percentile in the \lamostmcq sample, \jvs chose the 95th percentile. Similarly, our boundary for the CKS long-period pileup was chosen subjectively.}
\end{enumerate}

@@

\edit1{Finally, we note that we regard the short-period pileup as less secure than the long-period pileup. Preliminary tests with other density estimation methods \citep{Contardo2022}\footnote{\url{https://github.com/contardog/FindTheGap}} also reveal the short-period pileup, but further work is needed to better understand this feature. In Appendix~\ref{app:harmonics} we show that the short-period pileup does not appear to be a harmonic of the long-period pileup, as one might expect if the periods on the short-period pileup were erroneously determined. However, the short-period pileup is primarily observed in the \lamostmcq sample, while it is unclear whether it is present in the \lamostsan sample (see the rightmost panels of Figure~\ref{fig:xmatch}). The differences between the \mma and \santos catalogs may provide an explanation for this observation. The latter catalog contains a much larger number of detections at longer rotation periods, and the higher sensitivity to more slowly rotating stars within the \santos may then make the short-period pileup appear weaker relative to the long-period pileup. This is as expected since open cluster observations suggest that the short-period pileup can not be a long-lived feature, while the long-period pileup appears to be much longer lived. Consequently, the \santos catalog may more accurately reflect the relative strengths of these two features.}

@@

@@

    \item It is yet unclear whether the Sun resides on the long-period pileup or has already evolved through it. \edit1{Offsets of $\approx$110--140~K between the observational and theoretical \teff scales would place the long-period pileup at Rossby numbers consistent with other literature values of the location of the WMB transition.} If the Sun has evolved through the pileup, there is some modest tension with the Sun's age and the ages of the oldest stars on the pileup; some of the cooler long-period pileup stars in the CKS sample are both more massive and older than the Sun, which contradicts the expectation from the WMB model that hotter stars spend a shorter period of time on the pileup. However, this tension might be simply explained by inaccurate isochrone ages.
